---
title: "Single-cell ATAC-seq analysis: Public COVID1-9 ATAC-seq data by Wilk et al., JEM 2020"
author: "Jonas Schulte-Schrepping"
date: "2022.08.17"
output: 
  html_document: 
    code_download: yes
    df_print: kable
    theme: united
    toc: yes
    toc_depth: 8
    toc_float: yes
---

# Data sources:

https://rupress.org/jem/article-pdf/218/8/e20210582/1417996/jem_20210582.pdf
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE174072
https://github.com/ajwilk/COVID_scMultiome

Random seed
```{r}
set.seed(42)
```

#Global settings:

## Install packages

```{r}
# ArchR::installExtraPackages()
```

## Load packages

```{r, message=FALSE}
library(future)
library(future.apply)
library(tidyverse)
library(ggrepel)
library(rhdf5)
library(cowplot)
library(clusterProfiler)
library(BSgenome.Hsapiens.UCSC.hg38)
library(org.Hs.eg.db)
library(ChIPseeker)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(Seurat)
library(GenomeInfoDb)
library(EnsDb.Hsapiens.v86)
library(ggplot2)
library(patchwork)
library(reshape2)
library(SeuratDisk)
library(Cairo)
library(ArchR)
library(JASPAR2020)
library(UpSetR)
library(biomaRt)
```

## Settings for parallel computing
```{r}
# change the current plan to access seurat parallelization
plan("multiprocess", workers = 12)
options(future.globals.maxSize = 20000 * 1024^2)
addArchRThreads(threads = 1) 
```

## Load Hallmark Gene sets

```{r}
# transpose gene names to human symbols
human = useMart("ensembl", dataset = "hsapiens_gene_ensembl", host = "https://dec2021.archive.ensembl.org/")
mouse = useMart("ensembl", dataset = "mmusculus_gene_ensembl", host = "https://dec2021.archive.ensembl.org/")
```

```{r}
hallmark_genes <- clusterProfiler::read.gmt("h.all.v6.2.entrez.gmt")
```

## Functions 

### Confusion Matrix
```{r}
confusionMatrix <- function (i = NULL, j = NULL) 
{
  ui <- unique(i)
  uj <- unique(j)
  m <- Matrix::sparseMatrix(i = match(i, ui), j = match(j,uj), x = rep(1, length(i)), dims = c(length(ui), length(uj)))
  rownames(m) <- ui
  colnames(m) <- uj
  m
}
```


### Volcano Plot

```{r, fig.width=8, fig.height=6}
volcano_plot <- function(data,
                         useGroups,
                         labelnum = 20,
                         force_labels= NULL,
                         name_col = "name",
                         FDR_cutoff = 0.01,
                         FC_cutoff = 0.58,
                         x_lim=3){
  tmp <- getMarkers(data, cutOff = "FDR < 1")
  tmp <- as.data.frame(tmp[,useGroups])
  tmp$col <- ifelse(tmp$FDR <= FDR_cutoff & tmp$Log2FC >= FC_cutoff,"up",
                    ifelse(tmp$FDR <= FDR_cutoff & tmp$Log2FC <= -FC_cutoff,"down",
                           "n.s."))
  
  if(is.null(force_labels)){
    # specify labeling   
    upDE <-  as.data.frame(tmp[tmp$Log2FC >0,]) 
    FClabel_up <- upDE[order(abs(upDE$Log2FC), decreasing = TRUE),]
    if(nrow(FClabel_up)>labelnum){
      FClabel_up <- as.character(FClabel_up[c(1:labelnum),"name"])
    } else {
      FClabel_up <- as.character(FClabel_up[,"name"])}
    plabel_up <- upDE[order(upDE$FDR, decreasing = FALSE),]
    if(nrow(plabel_up)>labelnum){
      plabel_up <- as.character(plabel_up[c(1:labelnum),"name"])
    } else {
      plabel_up <- as.character(plabel_up[,gene_column])
    }
    
    # specify labeling   
    downDE <-  as.data.frame(tmp[tmp$Log2FC <0,]) 
    FClabel_down <- downDE[order(abs(downDE$Log2FC), decreasing = TRUE),]
    if(nrow(FClabel_down)>labelnum){
      FClabel_down <- as.character(FClabel_down[c(1:labelnum),"name"])
    } else {
      FClabel_down <- as.character(FClabel_down[,"name"])}
    plabel_down <- downDE[order(downDE$FDR, decreasing = FALSE),]
    if(nrow(plabel_down)>labelnum){
      plabel_down <- as.character(plabel_down[c(1:labelnum),"name"])
    } else {
      plabel_down <- as.character(plabel_down[,gene_column])
    }
    
    label<- unique(c(FClabel_up, plabel_up, FClabel_down, plabel_down))
    tmp$label<- ifelse(tmp$name %in% label == "TRUE",
                       as.character(tmp$name), "")
  }else{
    tmp$label <- ifelse(tmp$name %in% force_labels == "TRUE",
                       as.character(tmp$name), "")
  }
  
  p <- ggplot(data=tmp[!(tmp$Log2FC>-FC_cutoff & tmp$Log2FC< FC_cutoff | tmp$FDR > FDR_cutoff),], aes(x=Log2FC, y=-log10(FDR))) +
    scale_x_continuous(limits=c(-x_lim,x_lim))+
    geom_vline(xintercept = 0, colour="black")+
    geom_vline(xintercept = c(-0.58,0.58), colour="grey")+
    geom_hline(yintercept=-log(0.05,10),colour="grey")+
    geom_point(aes(colour=col), size=1) +
    xlab("log2(FoldChange)") +
    ylab("-log10(FDR)") +
    geom_text_repel(aes(label=label), size=2, max.overlaps = Inf, segment.size = 0.1)+
    geom_text(aes(x_lim-1, 100, label = sum(FDR <= FDR_cutoff &
                                         Log2FC >= FC_cutoff)), 
              col = "black")+
    geom_text(aes(-x_lim+1, 100, label = sum(FDR <= FDR_cutoff &
                                          Log2FC <= -FC_cutoff)), 
              col = "black")+
    theme_bw() + scale_color_manual(values=c("n.s."="grey", "up"="#a63232", "down"="#335599"))+ theme(legend.position = "none")
  
  p
}
```

### GSEA

```{r}
scGSEA <-  function(input,
                    pvalueCutoff = 0.1,
                    qvalueCutoff = 0.1){ 
  
  results <- list()
  
  # Open in Glucan
  symbol_up <- input[input$Log2FC>0,]$name
  entrez_up <- bitr(unique(symbol_up), 
                    fromType = "SYMBOL", 
                    toType="ENTREZID", 
                    OrgDb=org.Hs.eg.db)$ENTREZID
  symbol_down <- input[input$Log2FC<0,]$name
  entrez_down <- bitr(unique(symbol_down), 
                    fromType = "SYMBOL", 
                    toType="ENTREZID", 
                    OrgDb=org.Hs.eg.db)$ENTREZID
  # GO enrichment
    results$GOup <- as.data.frame(enrichGO(gene = entrez_up,
                                           universe = present_genes_entrez,
                                           OrgDb = org.Hs.eg.db,
                                           ont = "BP",
                                           pAdjustMethod = "bonferroni",
                                           pvalueCutoff  = pvalueCutoff,
                                           qvalueCutoff  = qvalueCutoff,
                                           readable      = T))
    
    results$GOdown <- as.data.frame(enrichGO(gene = entrez_down,
                                           universe = present_genes_entrez,
                                           OrgDb = org.Hs.eg.db,
                                           ont = "BP",
                                           pAdjustMethod = "bonferroni",
                                           pvalueCutoff  = pvalueCutoff,
                                           qvalueCutoff  = qvalueCutoff,
                                           readable      = T))
    
  # KEGG enrichment
  results$KEGGup <- as.data.frame(enrichKEGG(gene = entrez_up, 
                                               organism = "hsa",
                                               universe = present_genes_entrez,
                                               pAdjustMethod = "bonferroni",
                                               pvalueCutoff  = pvalueCutoff,
                                               qvalueCutoff = qvalueCutoff))
    
    results$KEGGdown <- as.data.frame(enrichKEGG(gene = entrez_down, 
                                               organism = "hsa",
                                               universe = present_genes_entrez,
                                               pAdjustMethod = "bonferroni",
                                               pvalueCutoff  = pvalueCutoff,
                                               qvalueCutoff = qvalueCutoff))
    
    # HALLMARK enrichment
    results$HALLMARKup <- as.data.frame(enricher(entrez_up,
                                                 TERM2GENE=hallmark_genes,
                                                 universe = present_genes_entrez,  
                                                 pAdjustMethod = "bonferroni",
                                                 pvalueCutoff  = pvalueCutoff,
                                                 qvalueCutoff  = qvalueCutoff))
    
    # HALLMARK enrichment
    results$HALLMARKdown <- as.data.frame(enricher(entrez_down,
                                                 TERM2GENE=hallmark_genes,
                                                 universe = present_genes_entrez,  
                                                 pAdjustMethod = "bonferroni",
                                                 pvalueCutoff  = pvalueCutoff,
                                                 qvalueCutoff  = qvalueCutoff))
    return(results)
}
```


### GSEA DotPlot

```{r}
dotplotGSEA <- function(x,
                        show=25,
                        font.size=10,
                        title=NULL,
                        title.size=10,
                        title.width=100,
                        order="count"){
  if(nrow(x)<1){
    print("No enrichment found.")
  }else{
    x <- if(nrow(x)>show){x[c(1:show),]}else{x}
    if(order=="padj"){
      x <- x[order(x$Count,decreasing=FALSE),]
      x$Description <- ifelse(nchar(x$Description)>50,
                              paste(substr(x$Description, 1, 50),"[...]",sep=""),
                              x$Description)
      x$GeneRatio <- factor(x$GeneRatio, levels = unique(x$GeneRatio))
      x <- x[order(x$p.adjust,decreasing=TRUE),]
      x$Description <- factor(x$Description, levels = unique(x$Description))
    }
    if(order=="count"){
      x <- x[order(x$Count,decreasing=FALSE),]
      x$Description <- ifelse(nchar(x$Description)>50,
                              paste(substr(x$Description, 1, 50),"[...]",sep=""),
                              x$Description)
      x$Description <- factor(x$Description, levels = unique(x$Description))
      x$GeneRatio <- factor(x$GeneRatio, levels = unique(x$GeneRatio))
    }
    ggplot(x, aes(x = Count, y = Description, color = p.adjust)) +
      geom_point(aes(size = Count)) +
      scale_colour_gradientn(colours=c('red', 
                                       'orange', 
                                       'darkblue',
                                       'darkblue'),
                             limits=c(0,1),
                             values   = c(0,0.05,0.2,0.5,1),
                             breaks   = c(0.05,0.2,1),
                             labels = format(c(0.05,0.2,1))) +
      ylab(NULL) +
      theme_bw() +
      theme(text = element_text(size=font.size),
            axis.text.x = element_text(angle = 90)) 
  }
}
```

### Get motif target regions

```{r}
getTargetRegion <- function(motif,
                            positions=motifPositions_homer,
                            query_GR,
                            peak_annotation){
  overlap <- findOverlaps(query = positions[[motif]], 
                          subject = query_GR, 
                          type = "within")
  target_region <- peak_annotation[unique(overlap@to),]
  target_region
}
```


# -------------------------------------------------------------------------------

# scATAC-seq data analysis using ArchR

Download data from Geo using GEOquery

```{r}
library("GEOquery")
```

```{r}
gse <- getGEO("GSE174072",GSEMatrix=FALSE)
head(Meta(gse))
names(GSMList(gse))
```

```{r}
for(i in 1:length(GSMList(gse))){
  print(names(GSMList(gse))[i])
  print(GSMList(gse)[[i]]@header$library_strategy)
}
```

```{r}
samples <- c("GSM5285728" , "GSM5285729", "GSM5285730", "GSM5285731", "GSM5285732",
             "GSM5285733", "GSM5285734", "GSM5285735", "GSM5285736", "GSM5285737",
             "GSM5285738", "GSM5285739", "GSM5285740","GSM5285741", "GSM5285742",
             "GSM5285743", "GSM5285744", "GSM5285745", "GSM5285746", "GSM5285747",
             "GSM5285748", "GSM5285749", "GSM5285750", "GSM5285751", "GSM5285752"
             )

for(i in samples){
  print(i)
  getGEOSuppFiles(GEO = i,
                baseDir = "Wilk_JEM_2021/ATAC",
                makeDirectory = TRUE)
}
```


```{r}
addArchRGenome("hg38")
```

## Create Arrow files

```{r}
inputFiles <- dir("Wilk_JEM_2021/ATAC/fragments", full.names = TRUE)
inputFiles <- inputFiles[grepl("_fragments.tsv.gz$", inputFiles)]
names <- dir("Wilk_JEM_2021/ATAC/fragments", full.names = FALSE)
names <- names[grepl("_fragments.tsv.gz$", names)]
names(inputFiles) <- unlist(lapply(strsplit(names, split = "_"), function(x) x[1]))

inputFiles
```

```{r}
ArrowFiles <- createArrowFiles(inputFiles = inputFiles,
                               sampleNames = names(inputFiles),
                               addTileMat = TRUE,
                               addGeneScoreMat = TRUE,
                               QCDir = "QualityControl_220531/"
)
```


## Create ArchRProject

```{r}
proj <- ArchRProject(
  ArrowFiles = ArrowFiles,
  outputDirectory = "ArchR_scATAC_Wilk",
  copyArrows = TRUE
)
```

```{r}
saveArchRProject(ArchRProj = proj, 
                 outputDirectory = "ArchR_scATAC_Wilk", 
                 overwrite = FALSE, 
                 load = TRUE)
```

Load previous Archr project 

```{r}
# proj <- loadArchRProject(path = "./ArchR_scATAC_Wilk/", force = FALSE, showLogo = TRUE)
```

#---

# Filter cells and add meta data

Match cell ids from ArchR project with cell meta information provided on github

```{r}
sample_match <- data.frame("GSM" = c("GSM5285728" , "GSM5285729", "GSM5285730", "GSM5285731", "GSM5285732",
                                     "GSM5285733", "GSM5285734", "GSM5285735", "GSM5285736", "GSM5285737",
                                     "GSM5285738", "GSM5285739", "GSM5285740","GSM5285741", "GSM5285742",
                                     "GSM5285743", "GSM5285744", "GSM5285745", "GSM5285746", "GSM5285747",
                                     "GSM5285748", "GSM5285749", "GSM5285750", "GSM5285751", "GSM5285752"),
                           "Sample"=c("ATAC_055", "ATAC_057", "ATAC_132D0", "ATAC_52", "ATAC_555_1", 
                                      "ATAC_555_2", "ATAC_556", "ATAC_557", "ATAC_558", "ATAC_559", 
                                      "ATAC_560", "ATAC_564A", "ATAC_564B", "ATAC_66D0", "ATAC_66D7",  
                                      "ATAC_67", "ATAC_83", "ATAC_86", "ATAC_EV08", "ATAC_HIP02_frozen", 
                                      "ATAC_HIP023_frozen", "ATAC_HIP043", "ATAC_HIP044", "ATAC_HIP045", "ATAC_HIP15_frozen"))

proj@cellColData$SampleID <- sample_match$Sample[match(as.character(proj@cellColData$Sample),
                                                       sample_match$GSM)]

proj@cellColData$Cell_id <- paste(proj@cellColData$SampleID, "#",
                                  unlist(lapply(strsplit(rownames(proj@cellColData), split = "#"), function(x) x[2])), sep="")
```

Meta data was downloaded from: https://github.com/ajwilk/COVID_scMultiome/blob/main/data/scATAC/atac_cell_metadata.csv

```{r}
meta <- read.csv("atac_cell_metadata.csv", stringsAsFactors = FALSE)
colnames(meta)
```

### Filter ArchR project

```{r}
proj_filt <- proj[proj$cellNames[which(proj$Cell_id %in% meta[!is.na(meta$current_severity_bin),]$cell_id)],]
proj_filt
```

### Add meta information

```{r}
idx <- match(as.character(proj_filt$Cell_id), meta$cell_id)

proj_filt$is_doublet <- meta$is_doublet[idx]
proj_filt$cell_type <- meta$cell_type[idx]
proj_filt$current_severity_bin <- meta$current_severity_bin[idx]
proj_filt$current_severity_bin <- ifelse(is.na(proj_filt$current_severity_bin),"n.d.",proj_filt$current_severity_bin)
```

We skip the doublet filtering, as we filtered the cells based on the information available in the cell meta annotation of the publication.

#--- 

# Quality Scores

```{r}
df <- getCellColData(proj_filt, select = c("log10(nFrags)", "TSSEnrichment"))

p <- ggPoint(
    x = df[,1], 
    y = df[,2], 
    colorDensity = TRUE,
    continuousSet = "sambaNight",
    xlabel = "Log10 Unique Fragments",
    ylabel = "TSS Enrichment",
    xlim = c(log10(500), quantile(df[,1], probs = 0.99)),
    ylim = c(0, quantile(df[,2], probs = 0.99))
) + geom_hline(yintercept = 4, lty = "dashed") + geom_vline(xintercept = 3, lty = "dashed")

p

rm(df, p)
```

### Plots for each sample for the TSS enrichment scores and log10(unique features)

```{r, fig.height=4, fig.width=12}
proj_filt$group_sample <- paste(proj_filt$current_severity_bin, proj_filt$SampleID)

p1 <- plotGroups(
    ArchRProj = proj_filt, 
    groupBy = "group_sample", 
    colorBy = "cellColData", 
    name = "TSSEnrichment",
    plotAs = "ridges"
   )

p2 <- plotGroups(
    ArchRProj = proj_filt, 
    groupBy = "group_sample", 
    colorBy = "cellColData", 
    name = "TSSEnrichment",
    plotAs = "violin",
    alpha = 0.4,
    addBoxPlot = TRUE
   )

p3 <- plotGroups(
    ArchRProj = proj_filt, 
    groupBy = "group_sample", 
    colorBy = "cellColData", 
    name = "log10(nFrags)",
    plotAs = "ridges"
   )

p4 <- plotGroups(
    ArchRProj = proj_filt, 
    groupBy = "group_sample", 
    colorBy = "cellColData", 
    name = "log10(nFrags)",
    plotAs = "violin",
    alpha = 0.4,
    addBoxPlot = TRUE
   )

ggAlignPlots(p1, p2, p3, p4, type = "h")
rm(p1,p2,p3,p4)
```

### Fragment Size distribution and TSS enrichment

```{r, fig.height3, fig.width=8}
p1 <- plotFragmentSizes(ArchRProj = proj_filt)
p2 <- plotTSSEnrichment(ArchRProj = proj_filt,
                        flank = 10000)

ggAlignPlots(p1,p2, type = "h")
rm(p1,p2)
```

#---

# Downstream analyses

## Dimensionality Reduction and Clustering

ArchR implements an iterative LSI dimensionality reduction via the addIterativeLSI() function.

```{r}
proj_filt <- addIterativeLSI(ArchRProj = proj_filt, useMatrix = "TileMatrix", name = "IterativeLSI")
```

To call clusters in this reduced dimension sub-space, we use the addClusters() function which uses Seurat’s graph clustering as the default clustering method.

```{r}
proj_filt <- addClusters(input = proj_filt,
                         reducedDims = "IterativeLSI",
                         method = "Seurat",
                         force = TRUE,
                         resolution = 0.8)
```

## Visualizing in a 2D UMAP Embedding

We can visualize our scATAC-seq data using a 2-dimensional representation such as Uniform Manifold Approximation and proj_filtection (UMAP). To do this, we add a UMAP embedding to our ArchRproj_filtect object with the addUMAP() function. This function uses the uwot package to perform UMAP.

```{r}
proj_filt <- addUMAP(ArchRProj = proj_filt, reducedDims = "IterativeLSI")
```

Using this UMAP, we can visualize various attributes of our cells which are stored in a matrix called cellColData in our ArchRproj_filtect. To do this, we use the plotEmbedding() function and we specify the variable to use for coloration via a combination of the colorBy and name parameters.

```{r}
plotEmbedding(ArchRProj = proj_filt, colorBy = "cellColData", 
              name = "cell_type",
              embedding = "UMAP", verbose=FALSE)
```

```{r}
plotEmbedding(ArchRProj = proj_filt, 
              colorBy = "cellColData", 
              name = "current_severity_bin", 
              pal = c("0"="#4f4d4c",
                      "1-3" ="#ffb867",
                      "4-5" = "#5D63A1",
                      "6-7" = "#982326"),
              embedding = "UMAP", verbose=FALSE)
```

#---

# Monocyte-specific analysis

```{r}
proj_mono<- proj_filt[proj_filt$cellNames[which(proj_filt$cell_type == "CD14 Mono")],]

proj_mono
```

### Dimensionality reduction

```{r}
proj_mono <- addIterativeLSI(ArchRProj = proj_mono, useMatrix = "GeneScoreMatrix", name = "IterativeLSI_mono", force = TRUE)
```

```{r}
proj_mono <- addUMAP(ArchRProj = proj_mono, reducedDims = "IterativeLSI_mono", name = "UMAP_mono")
```

```{r}
plotEmbedding(ArchRProj = proj_mono, colorBy = "cellColData", name = "current_severity_bin", 
              embedding = "UMAP_mono", verbose=FALSE)
```

#---

# Visualization of selected cluster marker genes {.tabset}

We add imputation weights using MAGIC to help smooth the dropout noise in our gene scores.

```{r}
proj_mono <- addImputeWeights(proj_mono)
```

```{r, fig.width=8, fig.height=3}
p1 <- plotGroups(
    ArchRProj = proj_mono,
    groupBy = "current_severity_bin", 
    colorBy = "GeneScoreMatrix", 
    name = "IL15",
    plotAs = "violin",
    pal = c("0"="#4f4d4c",
            "1-3" ="#ffb867",
            "4-5" = "#5D63A1",
            "6-7" = "#982326"),
    alpha = 0.4,
    addBoxPlot = TRUE
   )

p2 <- plotGroups(
    ArchRProj = proj_mono,
    groupBy = "current_severity_bin", 
    colorBy = "GeneScoreMatrix", 
    name = "CD83",
    plotAs = "violin",
    pal = c("0"="#4f4d4c",
            "1-3" ="#ffb867",
            "4-5" = "#5D63A1",
            "6-7" = "#982326"),
    alpha = 0.4,
    addBoxPlot = TRUE
   )

p3 <- plotGroups(
    ArchRProj = proj_mono,
    groupBy = "current_severity_bin", 
    colorBy = "GeneScoreMatrix", 
    name = "CXCL16",
    plotAs = "violin",
    pal = c("0"="#4f4d4c",
            "1-3" ="#ffb867",
            "4-5" = "#5D63A1",
            "6-7" = "#982326"),
    alpha = 0.4,
    addBoxPlot = TRUE
   )

p4 <- plotGroups(
    ArchRProj = proj_mono,
    groupBy = "current_severity_bin", 
    colorBy = "GeneScoreMatrix", 
    name = "TNF",
    plotAs = "violin",
    pal = c("0"="#4f4d4c",
            "1-3" ="#ffb867",
            "4-5" = "#5D63A1",
            "6-7" = "#982326"),
    alpha = 0.4,
    addBoxPlot = TRUE
   )

p <- CombinePlots(plots=list(p1,p2,p3,p4),ncol=4)

p
```

#---

# Gene level analysis

```{r}
GeneScoreMtx_mono <- getMatrixFromProject(
  ArchRProj = proj_mono,
  useMatrix = "GeneScoreMatrix"
)

present_genes_entrez <- bitr(GeneScoreMtx_mono@elementMetadata$name, 
                             fromType = "SYMBOL", 
                             toType="ENTREZID", 
                             OrgDb=org.Hs.eg.db)$ENTREZID
```


## Infer severity marker genes

```{r}
severity_markersGS <- getMarkerFeatures(
  ArchRProj = proj_mono, 
  useMatrix = "GeneScoreMatrix", 
  groupBy = "current_severity_bin",
  bias = c("TSSEnrichment", "log10(nFrags)"),
  testMethod = "wilcoxon"
)

severity_markerList <- getMarkers(severity_markersGS, cutOff = "FDR <= 0.05 & Log2FC >= 0.58")
```

### Heatmap of cluster marker genes

```{r}
heatmapGS <- plotMarkerHeatmap(
  seMarker = severity_markersGS, 
  cutOff = "FDR <= 0.01 & Log2FC >= 0.58", 
  labelMarkers = NULL,
  transpose = TRUE
)
```

```{r,fig.height=3.75, fig.width= 4}
ComplexHeatmap::draw(heatmapGS, heatmap_legend_side = "bot", annotation_legend_side = "bot")
```

## Pairwise test between severity groups

### Mild-specific genes

```{r}
markerGenes_mild <- getMarkerFeatures(
  ArchRProj = proj_mono, 
  useMatrix = "GeneScoreMatrix",
  groupBy = "current_severity_bin",
  testMethod = "binomial",
  maxCells=2000,
  bias = c("TSSEnrichment", "log10(nFrags)"),
  useGroups = "1-3",
  bgdGroups = "0", 
  binarize=TRUE
)
```

#### Volcano Plot

```{r,fig.width=5,fig.height=4}
p <- volcano_plot(data=markerGenes_mild,
                  useGroups="1-3",
                  labelnum = 20,
                  name_col = "name",
                  FDR_cutoff = 0.05,
                  FC_cutoff = 0.58)

p
```

#### GSEA

```{r,fig.width=5, fig.height=5}
markerList_mild <- getMarkers(markerGenes_mild, 
                             cutOff = "FDR <= 0.05 & abs(Log2FC) >= 0.58")

GSEA_genes_mild <- scGSEA(input = markerList_mild$`1-3`,
                          pvalueCutoff = 0.1,
                          qvalueCutoff = 0.1)

dotplotGSEA(x=GSEA_genes_mild$HALLMARKup)
```

#------------------------------------------------------------------------------

# Peak level analysis

## Pseudobulk of samples

```{r}
proj_mono <- addGroupCoverages(ArchRProj = proj_mono,
                               groupBy = "current_severity_bin")
```

## MACS3 peak calling


```{r}
proj_mono <- addReproduciblePeakSet(
   ArchRProj = proj_mono,
   groupBy = "current_severity_bin",
   reproducibility = "3",
   pathToMacs2 = "macs3",
   minCells = 100,
   verbose=FALSE
)
 
allPeaks <- getPeakSet(proj_mono)
```

Add peak matrix to project:

```{r}
proj_mono <- addPeakMatrix(proj_mono)
```

### Mild-specific peaks

```{r}
markerPeaks_mild <- getMarkerFeatures(
  ArchRProj = proj_mono, 
  useMatrix = "PeakMatrix",
  groupBy = "current_severity_bin",
  testMethod = "binomial",
  maxCells=2000,
  bias = c("TSSEnrichment", "log10(nFrags)"),
  useGroups = "1-3",
  bgdGroups = "0", 
  binarize=TRUE
)
```

```{r}
markerList_mild_up <- getMarkers(markerPeaks_mild, 
                                 cutOff = "FDR <= 0.01 & Log2FC >= 2", 
                                 returnGR = TRUE)
```

```{r}
peakAnnotation_markerList_mild_up <- annotatePeak(markerList_mild_up$`1-3`,
                                              TxDb = TxDb.Hsapiens.UCSC.hg38.knownGene, 
                                              level="gene", 
                                              annoDb="org.Hs.eg.db")

peakAnnotation_markerList_mild_up.df <- as.data.frame(as.GRanges(peakAnnotation_markerList_mild_up))

rownames(peakAnnotation_markerList_mild_up.df) <- paste(peakAnnotation_markerList_mild_up.df$seqnames,
                                                     peakAnnotation_markerList_mild_up.df$start,
                                                     peakAnnotation_markerList_mild_up.df$end, sep="_")
```

### Severe-specific peaks

```{r}
markerPeaks_severe <- getMarkerFeatures(
  ArchRProj = proj_mono, 
  useMatrix = "PeakMatrix",
  groupBy = "current_severity_bin",
  testMethod = "binomial",
  maxCells=2000,
  bias = c("TSSEnrichment", "log10(nFrags)"),
  useGroups = "6-7",
  bgdGroups = "0", 
  binarize=TRUE
)
```

```{r}
markerList_severe_up <- getMarkers(markerPeaks_severe, 
                                 cutOff = "FDR <= 0.01 & Log2FC >= 2", 
                                 returnGR = TRUE)
```

```{r}
peakAnnotation_markerList_severe_up <- annotatePeak(markerList_severe_up$`6-7`,
                                              TxDb = TxDb.Hsapiens.UCSC.hg38.knownGene, 
                                              level="gene", 
                                              annoDb="org.Hs.eg.db")

peakAnnotation_markerList_severe_up.df <- as.data.frame(as.GRanges(peakAnnotation_markerList_severe_up))

rownames(peakAnnotation_markerList_severe_up.df) <- paste(peakAnnotation_markerList_severe_up.df$seqnames,
                                                     peakAnnotation_markerList_severe_up.df$start,
                                                     peakAnnotation_markerList_severe_up.df$end, sep="_")
```


# ----------------------------------------------------

# ArchR TF binding motif enrichment analysis

```{r}
proj_mono <- addMotifAnnotations(ArchRProj = proj_mono, motifSet = "homer", name = "Motif_homer",force=TRUE)
```

```{r}
motifPositions_homer <- getPositions(proj_mono, 
                                     name = "Motif_homer")

motifPositions_homer
```

## Mild COVID19 OPEN DAR

### HOMER Motif Analysis

```{r}
motifsUp_mild <- peakAnnoEnrichment(
    seMarker = markerPeaks_mild,
    ArchRProj = proj_mono,
    background = "all",
    peakAnnotation = "Motif_homer",
    cutOff = "FDR <= 0.01 & Log2FC >= 2"
  )


motifsUp_mild_homer <- data.frame(TF = rownames(motifsUp_mild), mlog10Padj = assay(motifsUp_mild)[,1])
motifsUp_mild_homer <- motifsUp_mild_homer[order(motifsUp_mild_homer$mlog10Padj, decreasing = TRUE),]
motifsUp_mild_homer$rank <- as.numeric(seq_len(nrow(motifsUp_mild_homer)))
```

```{r}
target_list <- list()
for(i in motifsUp_mild_homer$TF){
  print(i)
  tmp  <- getTargetRegion(motif=i,
                          positions=motifPositions_homer,
                          query_GR=getMarkers(markerPeaks_mild, 
                                              cutOff = "FDR <= 0.01 & Log2FC >= 2", 
                                              returnGR = TRUE)$`1-3`,
                          peak_annotation=peakAnnotation_markerList_mild_up.df)
  target_list[[paste(i)]] <- tmp
  #target_list[[paste(i)]] <- tmp[abs(tmp$distanceToTSS) < 10000,]
}

motifsUp_mild_homer$n_targets <- sapply(target_list, nrow)

motifsUp_mild_homer <- motifsUp_mild_homer[order(motifsUp_mild_homer$n_targets, decreasing = TRUE),]
motifsUp_mild_homer$nrank <- as.numeric(seq_len(nrow(motifsUp_mild_homer)))
```


```{r, fig.width=12, fig.height=12}
tmp <- motifsUp_mild_homer[order(motifsUp_mild_homer$rank, decreasing = FALSE),]
top50_rank <- tmp[c(1:50),]

p1 <- ggplot(top50_rank, aes(x = rank, y = mlog10Padj, color = mlog10Padj, size= n_targets)) + 
  geom_point() +
  scale_y_continuous(limits=c(0,270))+
  ggrepel::geom_text_repel(
        data = top50_rank, 
        aes(x = rank, y = mlog10Padj, label = TF), 
        size = 2,
        force_pull   = 0, # do not pull toward data points
        nudge_y      = 0.05,
        direction    = "x",
        angle        = 90,
        hjust        = -0.5,
        segment.size = 0.2,
        max.iter = 1e4, max.time = 1,
        color = "black",
        max.overlaps = Inf
  ) + 
  theme_ArchR() + 
  ylab("-log10(P-adj) Motif Enrichment") + 
  xlab("p-value Rank Sorted TFs Enriched") +
  scale_color_gradientn(colors = paletteContinuous(set = "comet"))+
  ggtitle("mild (homer motifs)")

tmp <- motifsUp_mild_homer[order(motifsUp_mild_homer$nrank),]
top50_nrank <- tmp[c(1:50),]

p2 <- ggplot(top50_nrank, aes(x = nrank, y = n_targets, color = mlog10Padj, size= n_targets)) + 
  geom_point() +
  scale_y_continuous(limits=c(0,1200))+
  ggrepel::geom_text_repel(
        data = top50_nrank, 
        aes(x = nrank, y = n_targets, label = TF), 
        size = 2,
        force_pull   = 0, # do not pull toward data points
        nudge_y      = 0.05,
        direction    = "x",
        angle        = 90,
        hjust        = -0.5,
        segment.size = 0.2,
        max.iter = 1e4, max.time = 1,
        color = "black",
        max.overlaps = Inf
  ) + theme_ArchR() + 
  ylab("Number of target regions") + 
  xlab("n target Rank Sorted TFs Enriched") +
  scale_color_gradientn(colors = paletteContinuous(set = "comet"))+
  ggtitle("mild (homer motifs)")

p1/p2
```

### Upset Plot of target regions of key motifs

```{r,fig.width=8,fig.height=4}
target_regions_list <- list("Fra1.bZIP_98" = rownames(target_list[["Fra1.bZIP_98"]]),
                    "Fosl2.bZIP_86" = rownames(target_list[["Fosl2.bZIP_86"]]),
                    "Atf3.bZIP_12" = rownames(target_list[["Atf3.bZIP_12"]]),
                    "Fra2.bZIP_99" =rownames(target_list[["Fra2.bZIP_99"]]),
                    "Jun.AP1.bZIP_141" = rownames(target_list[["Jun.AP1.bZIP_141"]]),
                    "AP.1.bZIP_1" = rownames(target_list[["AP.1.bZIP_1"]]),
                    "BATF.bZIP_20" = rownames(target_list[["BATF.bZIP_20"]]),
                    "Bach2.bZIP_17" = rownames(target_list[["Bach2.bZIP_17"]]),
                    "Bach1.bZIP_16" = rownames(target_list[["Bach1.bZIP_16"]]),
                    "NF.E2.bZIP_179" = rownames(target_list[["NF.E2.bZIP_179"]]),
                    "IRF1.IRF_136" = rownames(target_list[["IRF1.IRF_136"]]),
                    "IRF.BATF.IRF.bZIP_19" = rownames(target_list[["IRF.BATF.IRF.bZIP_19"]]),
                    "NFkB.p65.Rel.RHD_180" = rownames(target_list[["NFkB.p65.Rel.RHD_180"]])
                    )

p <- upset(fromList(target_regions_list), order.by = "freq", nsets = 13, mb.ratio = c(0.4, 0.6))
p
```

## Severe COVID19 OPEN DAR

### HOMER Motif Analysis

```{r}
motifsUp_severe <- peakAnnoEnrichment(
    seMarker = markerPeaks_severe,
    ArchRProj = proj_mono,
    background = "all",
    peakAnnotation = "Motif_homer",
    cutOff = "FDR <= 0.01 & Log2FC >= 2"
  )


motifsUp_severe_homer <- data.frame(TF = rownames(motifsUp_severe), mlog10Padj = assay(motifsUp_severe)[,1])
motifsUp_severe_homer <- motifsUp_severe_homer[order(motifsUp_severe_homer$mlog10Padj, decreasing = TRUE),]
motifsUp_severe_homer$rank <- as.numeric(seq_len(nrow(motifsUp_severe_homer)))
```

```{r}
target_list_severe <- list()
for(i in motifsUp_severe_homer$TF){
  print(i)
  tmp  <- getTargetRegion(motif=i,
                          positions=motifPositions_homer,
                          query_GR=getMarkers(markerPeaks_severe, 
                                              cutOff = "FDR <= 0.01 & Log2FC >= 2", 
                                              returnGR = TRUE)$`6-7`,
                          peak_annotation=peakAnnotation_markerList_severe_up.df)
  target_list_severe[[paste(i)]] <- tmp
  #target_list_severe[[paste(i)]] <- tmp[abs(tmp$distanceToTSS) < 10000,]
}

motifsUp_severe_homer$n_targets <- sapply(target_list_severe, nrow)

motifsUp_severe_homer <- motifsUp_severe_homer[order(motifsUp_severe_homer$n_targets, decreasing = TRUE),]
motifsUp_severe_homer$nrank <- as.numeric(seq_len(nrow(motifsUp_severe_homer)))
```


```{r, fig.width=12, fig.height=12}
tmp <- motifsUp_severe_homer[order(motifsUp_severe_homer$rank, decreasing = FALSE),]
top50_rank <- tmp[c(1:50),]

p1 <- ggplot(top50_rank, aes(x = rank, y = mlog10Padj, color = mlog10Padj, size= n_targets)) + 
  geom_point() +
  scale_y_continuous(limits=c(0,270))+
  ggrepel::geom_text_repel(
        data = top50_rank, 
        aes(x = rank, y = mlog10Padj, label = TF), 
        size = 2,
        force_pull   = 0, # do not pull toward data points
        nudge_y      = 0.05,
        direction    = "x",
        angle        = 90,
        hjust        = -0.5,
        segment.size = 0.2,
        max.iter = 1e4, max.time = 1,
        color = "black",
        max.overlaps = Inf
  ) + 
  theme_ArchR() + 
  ylab("-log10(P-adj) Motif Enrichment") + 
  xlab("p-value Rank Sorted TFs Enriched") +
  scale_color_gradientn(colors = paletteContinuous(set = "comet"))+
  ggtitle("severe (homer motifs)")

tmp <- motifsUp_severe_homer[order(motifsUp_severe_homer$nrank),]
top50_nrank <- tmp[c(1:50),]

p2 <- ggplot(top50_nrank, aes(x = nrank, y = n_targets, color = mlog10Padj, size= n_targets)) + 
  geom_point() +
  scale_y_continuous(limits=c(0,1200))+
  ggrepel::geom_text_repel(
        data = top50_nrank, 
        aes(x = nrank, y = n_targets, label = TF), 
        size = 2,
        force_pull   = 0, # do not pull toward data points
        nudge_y      = 0.05,
        direction    = "x",
        angle        = 90,
        hjust        = -0.5,
        segment.size = 0.2,
        max.iter = 1e4, max.time = 1,
        color = "black",
        max.overlaps = Inf
  ) + theme_ArchR() + 
  ylab("Number of target regions") + 
  xlab("n target Rank Sorted TFs Enriched") +
  scale_color_gradientn(colors = paletteContinuous(set = "comet"))+
  ggtitle("severe (homer motifs)")

p1/p2
```

### Upset Plot of target regions of key motifs

```{r}
target_regions_list_severe <- list("Fra1.bZIP_98" = rownames(target_list_severe[["Fra1.bZIP_98"]]),
                                   "Fra2.bZIP_99" =rownames(target_list_severe[["Fra2.bZIP_99"]]),
                                   "Fosl2.bZIP_86" = rownames(target_list_severe[["Fosl2.bZIP_86"]]),
                                   "Atf3.bZIP_12" = rownames(target_list_severe[["Atf3.bZIP_12"]]),
                                   "AP.1.bZIP_1" = rownames(target_list_severe[["AP.1.bZIP_1"]]),
                                   "BATF.bZIP_20" = rownames(target_list_severe[["BATF.bZIP_20"]]),
                                   "Jun.AP1.bZIP_141" = rownames(target_list_severe[["Jun.AP1.bZIP_141"]]),
                                   "Bach2.bZIP_17" = rownames(target_list_severe[["Bach2.bZIP_17"]]),
                                   "Bach1.bZIP_16" = rownames(target_list_severe[["Bach1.bZIP_16"]]),
                                   "NF.E2.bZIP_179" = rownames(target_list_severe[["NF.E2.bZIP_179"]]),
                                   "IRF1.IRF_136" = rownames(target_list_severe[["IRF1.IRF_136"]]),
                                   "IRF.BATF.IRF.bZIP_19" = rownames(target_list_severe[["IRF.BATF.IRF.bZIP_19"]]),
                                   "NFkB.p65.Rel.RHD_180" = rownames(target_list_severe[["NFkB.p65.Rel.RHD_180"]])
)

p <- upset(fromList(target_regions_list_severe), order.by = "freq", nsets = 13, mb.ratio = c(0.4, 0.6))
p
```

# ----------------------------------------------------

# Clean up

```{r}
Sys.Date()
sessionInfo()
```
